---
title: "edx_flora_movielens_recommandation_system_project"
author: "Flora NIYOKWIZERWA"
date: "11/18/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction
the movie lens recommendation system project is part of HarvardX: PH125.9x Data Science: Capstone course.With this project I created movie recommendation system using movieLens dataset. I used 10M version of movieLens dataset. I downloaded the  dataset using the guidelines given from the course.
The goal of the project was to train machine learning algorithm and predict the movie rating in the validation set.
the steps followed: we downloaded the 10M movieLens dataset,train algorithm, analyse data, data visualization and used RMSE.

## we downloaded dataset from MovieLens as per instruction from the course 
the purpose of this part is to understand the use of dataset and help me to work on our project and familiarize myself with libraries to be used

```{r,echo=TRUE,message=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip
filename <- "comeon.R"

  dl <- tempfile()
  download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
  
  ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                   col.names = c("userId", "movieId", "rating", "timestamp"))
  
  movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
  colnames(movies) <- c("movieId", "title", "genres")
  

  # if using R 4.0 or later:
  movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                             title = as.character(title),
                                             genres = as.character(genres))
  
  
  movielens <- left_join(ratings, movies, by = "movieId")
  
  # Validation set will be 10% of MovieLens data
  set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
  test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
  edx <- movielens[-test_index,]
  temp <- movielens[test_index,]
  
  # Make sure userId and movieId in validation set are also in edx set
  validation <- temp %>% 
    semi_join(edx, by = "movieId") %>%
    semi_join(edx, by = "userId")
  
  # Add rows removed from validation set back into edx set
  removed <- anti_join(temp, validation)
  edx <- rbind(edx, removed)
  
  rm(dl, ratings, movies, test_index, temp, movielens, removed)
  
  #glimpse(edx)
  #we want to check the exact year movie was rated
  #so we convert timestamp to year
  edx <- mutate(edx,year_rated= year(as_datetime(timestamp)))
  head(edx)
  
 #data cleaning
  #we want to see the number of unique movies we have without repetition
  n_distinct(edx$movieId)
  #we also need to see number of unique users using userId who rated our movies
  n_distinct(edx$userId)
  #we start exploring our data set
  #firstly let's take a look on our data structure using str function
  str(edx)
  #now let's see our movies summary which presents statistical summary of our data set
  summary(edx)
  #we need to see the most watched movies from our data set and plot the result on bar chart
  edx %>%
    group_by(title) %>%
    summarize(count = n()) %>%
    arrange(-count) %>%
    top_n(10, count) %>%
    ggplot(aes(count, reorder(title, count))) +
    geom_bar(color = "black", fill = "deepskyblue2", stat = "identity") +
    xlab("Count") +
    ylab(NULL) +
    theme_bw()
  
  #the bar graph to show the highest rate given to any movie by any user
  edx %>%
    ggplot(aes(rating, y = ..prop..)) +
    geom_bar(color = "black", fill = "deepskyblue2") +
    labs(x = "Ratings", y = "Relative Frequency") +
    scale_x_continuous(breaks = c(0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5)) +
    theme_bw()
```

##Methods/analysis section
Within this section,I'm going to discuss different models used to analyse my dataset.I'm going to use different model and the residual mean square error(RMSE) to measure the accuracy and quantify typical error I may make when predicting the movie rating.
## the first model used is Simple mean Rating.
this method uses the mean of whole dataset to predict movie rating by assuming that any difference is due to random error.

```{r,echo=TRUE}
  aver <- mean(edx$rating)
  aver
  simplermse<-RMSE(validation$rating,aver)
  simplermse
  ## save the result from rmse using simple mean into data frame
  simple_rmse = data_frame(method= "Simple Mean model",
                           RMSE=simplermse)
  simple_rmse %>% knitr::kable()

```

the above model produced RMSE=1.061202 which is above 0.90000 and this typical error is larger than accepted error in our project

## the second model used is Movie Bias Effect model
after analysing with the simple mean rating model I found that error is large than accepted one, so I conducted further analysis to come up with movie rating prediction with small error
The movie bias effect model it uses the mean from the dataset and taking consideration of movie bias b_i based on difference between movie mean rating and overall rating

```{r,echo=TRUE}
#calculate overall dataset mean rating and store it in variable aver
  aver <- mean(edx$rating)
  aver
  #calculate movie rating mean
  movies_aver <- edx %>% group_by(movieId) %>% summarise(b_i= mean(rating-aver))
  p_rating <- aver + validation %>%
    left_join(movies_aver, by="movieId") %>%
    pull(b_i)
  rmse_modelmoviebias <- RMSE(p_rating, validation$rating)
  rmse_modelmoviebias
  simple_rmse <- bind_rows(simple_rmse, tibble(Method = "movie bias model", 
                                               RMSE=rmse_modelmoviebias))
  simple_rmse %>% knitr::kable()
  
```

with the above approach, it gave me lower error compared to simple mean rating model, the rmse got: 0.94390. But still the error is elevated so I had to continue my analysis with other approach.

## the third model used was movies and users effect model
with this model, I assumed that user feelings may affect movie rating outcome, that's why i used this model taking into consideration user mean rating b_u

```{r,echo=TRUE}
user_aver <- edx %>%
    left_join(movies_aver, by="movieId") %>%
    group_by(userId)%>%
    summarise(b_u = mean(rating - aver - b_i))
p_rating <- validation %>%
  left_join(movies_aver, by="movieId")%>%
  left_join(user_aver, by="userId") %>%
  mutate(pred = aver + b_i + b_u)%>%
  pull(pred)

rmse_user_effect <- RMSE(p_rating,validation$rating)
rmse_user_effect
simple_rmse <- bind_rows(simple_rmse,
                         data_frame(method="Movie-User effects",
                                    RMSE=rmse_user_effect))
simple_rmse %>% knitr::kable()
```

the above method produced low error of 0.8653488 compared to 0.9439087 produced by movie bias effect mode, I continued my analysis to the next model

##Regularization model

Regularisation allows for reduced errors caused by movies with few ratings which can influence the prediction and skew the error metric. The method uses a tuning parmeter, $\lambda$, to minimise the RMSE. Modifying $b_i$ and $b_u$ for movies with limited ratings.

```{r,echo=TRUE}
lambdas <- seq(0,5,0.5)
reguralizationrmses <- sapply(lambdas, function(x){
  aver <- mean(edx$rating)
  
  b_i <- edx %>%
    group_by(movieId) %>%
    summarise(b_i = sum(rating - aver)/(n()+x))
  b_u <- edx %>%
    left_join(b_i, by="movieId")%>%
    group_by(userId)%>%
    summarise(b_u = sum(rating - b_i - aver)/(n()+x))
  p_rating <- validation %>%
    left_join(b_i, by ="movieId")%>%
    left_join(b_u, by ="userId") %>%
    mutate(pred = aver + b_i + b_u)%>%
    pull(pred)
  
  return(RMSE(p_rating, validation$rating))
})
rmes_regul <- min(reguralizationrmses)
rmes_regul

#we plot RMSE against lambdas

qplot(lambdas,reguralizationrmses)
lambda<- lambdas[which.min(reguralizationrmses)]
lambda

simple_rmse<-bind_rows(simple_rmse,
                       data_frame(method="Regularization model",
                                  RMSE= rmes_regul))
simple_rmse %>% knitr::kable()

```

with this method I got rmse: 0.86418177 which is somehow the nearest accepted typical error.


##Results

the results of the models above can be found in our data frame table below
the lowest rmse found is 0.8648177

```{r,echo=TRUE}
simple_rmse %>% knitr::kable()
```

##Conclusion

The project objective was to create movie recommendation system using movieLens data set version 10M to predict the movie ratings.
I used different models to construct this project and got the lowest Residual mean square Error (RMSE).
The RMSE got was 0.8648177.
